Inferring the gravitational potential of the Milky Way with a few precisely measured stars

### 2014-04-28 text/paper2/ms.tex
start describing tub marg

To model the true, 6D, present-day position of the star, $\W$, we first transform to cartesian, Galactocentric coordinates where ($\bs{r}_0,\bs{v}_0$) and ($\bs{r}_{\sat,0},\bs{v}_{\sat,0}$) are the position and velocity of the star and progenitor today. The star is taken as having been sampled at $t=\tub$ (the unbinding time) from an isotropic Gaussian centered on one of the effective Lagrange points in position, and an isotropic Gaussian centered on the origin in velocity. The present-day phase-space position is the result of integrating this sample forward from $\tub$ in the parent potential, $\Phi$, whose form is parametrized by the vector $\bs{\Theta}_\Phi$. We assume that once the star becomes ``unbound'' from the satellite the potential of the satellite can be ignored,\footnote{This assumption breaks down for sufficiently large-mass progenitors, somewhere between $\sim10^8-10^9~\msun$.} and thus treat the star as a test particle. The satellite mass enters this prescription through the tidal radius; a changing satellite mass will skew the positions of the effective Lagrange points and velocity scale. We allow for mass-loss by incorporating the initial satellite mass, $m_0$, and a constant mass-loss term, $\dot{m}$ into the model: $m(t) = m_0 - \dot{m}t$. Finally, we add two additional parameters: a constant, global factor, $\Loffset$, that scales the position of the effective Lagrange points relative to the classical tidal radius, and a binary parameter $\tailbit$, that is either $-1$ or $+1$ depending on whether the star is in the leading or trailing tail. To compress notation, we pack all progenitor parameters ($\W_\sat, m_0, \dot{m}, \Loffset$) into the vector $\bs{\Theta}_\sat$, and all star parameters ($\W, \tailbit, \tub$) into the vector $\bs{\Theta}$. The likelihood is then:
\begin{align}
	p(\bs{\Theta} \given \bs{\Theta}_\sat, \bs{\Theta}_\Phi) &= p(\bs{r}_0,\bs{v}_0, \given \bs{\Theta}_\sat, \bs{\Theta}_\Phi)\,\left\vert\J\right\vert_{t=\tub}\label{eq:starlike}\\
	p(\bs{r}_0,\bs{v}_0, \given \bs{\Theta}_\sat, \bs{\Theta}_\Phi) &= \mathcal{N}(\bs{r} \given \bs{r}_\sat - \Loffset\,\tailbit\,\rtide\,\hat{x}_1, \rtide^2)\,\mathcal{N}(\bs{v} \given \bs{v}_\sat, \sigma^2_v)_{t=\tub}
\end{align}
where $\rtide$ is given by Eq.~\ref{eq:tidalradius}, and $\sigma_v$ by Eq.~\ref{eq:velscale}. $\left\vert\J\right\vert = \left\vert\frac{\partial(x,y,z,v_x,v_y,v_z)}{\partial(l,b,d,\mu_l,\mu_b,v_r)}\right\vert$ is the absolute value of the determinant of the Jacobian that defines the transformation from heliocentric, spherical to Galactocentric, cartesian coordinates. All quantities in this expression are evaluated at $t=\tub$. For a single star, the evaluation of the likelihood works as follows: given present-day coordinates for the star and progenitor, potential parameter values ($\bs{\Theta}_\Phi$), and nuisance parameter values ($\tub,\Loffset,\tailbit$),
\begin{enumerate}
	\item transform star and progenitor positions in heliocentric coordinates to Galactocentric, cartesian coordinates;
	\item integrate star and progenitor orbits backwards as test particles in the potential given by $\bs{\Theta}_\Phi$ for the total interaction time, $t_{\rm int}\approx6$~Gyr;
	\item at the timestep closest to $\tub$, transform the particle position to the relative, normalized coordinates defined in Section~\ref{sec:debris};
	\item compute the likelihood given by Equation~\ref{eq:starlike}.
\end{enumerate}
Assuming each star is an independent tracer, the full likelihood for many stars is then just the product of the individual likelihoods:
%\begin{multline}
%	p(\{\bs{\Theta}^{(i)}\}, \bs{\Theta}_\sat, \bs{\Theta}_\Phi \given \{\D^{(i)}\}, \D_\sat) = \\
%		\frac{1}{\mathcal{Z}} \, \prod_i \, p(\D^{(i)} \given \W^{(i)}) \, p(\D_\sat \given \W_\sat)\,
%			p(\bs{\Theta}^{(i)} \given \bs{\Theta}_\sat, \bs{\Theta}_\Phi) \, 
%			p(\bs{\Theta}) \, p(\bs{\Theta}_\sat)\,p(\bs{\Theta}_\Phi)
%\end{multline}
\begin{multline}
	p(\{\D^{(i)}\}, \D_\sat \given \{\bs{\Theta}^{(i)}\}, \bs{\Theta}_\sat, \bs{\Theta}_\Phi ) = \\
		p(\D_\sat \given \W_\sat) \, \prod_i \, p(\D^{(i)} \given \W^{(i)}) \,
			p(\bs{\Theta}^{(i)} \given \bs{\Theta}_\sat, \bs{\Theta}_\Phi)
\end{multline}
All parameters are summarized in Table~\ref{tbl:params}.

>>>>

To model the true, 6D, present-day position of the star, $\W$, we first transform to cartesian, Galactocentric coordinates where ($\bs{r}_0,\bs{v}_0$) and ($\bs{r}_{\sat,0},\bs{v}_{\sat,0}$) are the position and velocity of the star and progenitor today. The star is taken as having been sampled at $t=\tub$ (the unbinding time) from an isotropic Gaussian centered on one of the effective Lagrange points in position, and an isotropic Gaussian centered on the origin in velocity. The present-day phase-space position is the result of integrating this sample forward from $\tub$ in the parent potential, $\Phi$, whose form is parametrized by the vector $\bs{\Theta}_\Phi$. We assume that once the star becomes ``unbound'' from the satellite the potential of the satellite can be ignored,\footnote{This assumption breaks down for sufficiently large-mass progenitors, somewhere between $\sim10^8-10^9~\msun$.} and thus treat the star as a test particle. The satellite mass enters this prescription through the tidal radius; a changing satellite mass will skew the positions of the effective Lagrange points and velocity scale. We allow for mass-loss by incorporating the initial satellite mass, $m_0$, and a constant mass-loss term, $\dot{m}$ into the model: $m(t) = m_0 - \dot{m}t$. Finally, we add two additional parameters: a constant, global factor, $\Loffset$, that scales the position of the effective Lagrange points relative to the classical tidal radius, and a binary parameter $\tailbit$, that is either $-1$ or $+1$ depending on whether the star is in the leading or trailing tail. To compress notation, we pack all progenitor parameters ($\W_\sat, m_0, \dot{m}, \Loffset$) into the vector $\bs{\Theta}_\sat$, and all star parameters ($\W, \tailbit, \tub$) into the vector $\bs{\Theta}$. The likelihood is then:
\begin{align}
	p(\bs{\Theta} \given \bs{\Theta}_\sat, \bs{\Theta}_\Phi) &= p(\bs{r}_0,\bs{v}_0, \given \bs{\Theta}_\sat, \bs{\Theta}_\Phi)\,\left\vert\J\right\vert_{t=\tub}\label{eq:starlike}\\
	p(\bs{r}_0,\bs{v}_0, \given \bs{\Theta}_\sat, \bs{\Theta}_\Phi) &= \mathcal{N}(\bs{r} \given \bs{r}_\sat - \Loffset\,\tailbit\,\rtide\,\hat{x}_1, \rtide^2)\,\mathcal{N}(\bs{v} \given \bs{v}_\sat, \sigma^2_v)_{t=\tub}
\end{align}
where $\rtide$ is given by Eq.~\ref{eq:tidalradius}, and $\sigma_v$ by Eq.~\ref{eq:velscale}. $\left\vert\J\right\vert = \left\vert\frac{\partial(x,y,z,v_x,v_y,v_z)}{\partial(l,b,d,\mu_l,\mu_b,v_r)}\right\vert$ is the absolute value of the determinant of the Jacobian that defines the transformation from heliocentric, spherical to Galactocentric, cartesian coordinates. All quantities in this expression are evaluated at $t=\tub$.

We have found that leaving the unbinding time as a free parameter makes the likelihood severely multi-modal.

For a single star, the evaluation of the likelihood works as follows: given present-day coordinates for the star and progenitor, potential parameter values ($\bs{\Theta}_\Phi$), and nuisance parameter values ($\tub,\Loffset,\tailbit$),
\begin{enumerate}
	\item transform star and progenitor positions in heliocentric coordinates to Galactocentric, cartesian coordinates;
	\item integrate star and progenitor orbits backwards as test particles in the potential given by $\bs{\Theta}_\Phi$ for the total interaction time, $t_{\rm int}\approx6$~Gyr;
	\item at the timestep closest to $\tub$, transform the particle position to the relative, normalized coordinates defined in Section~\ref{sec:debris};
	\item compute the likelihood given by Equation~\ref{eq:starlike}.
\end{enumerate}

Assuming each star is an independent tracer, the full likelihood for many stars is then just the product of the individual likelihoods:
%\begin{multline}
%	p(\{\bs{\Theta}^{(i)}\}, \bs{\Theta}_\sat, \bs{\Theta}_\Phi \given \{\D^{(i)}\}, \D_\sat) = \\
%		\frac{1}{\mathcal{Z}} \, \prod_i \, p(\D^{(i)} \given \W^{(i)}) \, p(\D_\sat \given \W_\sat)\,
%			p(\bs{\Theta}^{(i)} \given \bs{\Theta}_\sat, \bs{\Theta}_\Phi) \, 
%			p(\bs{\Theta}) \, p(\bs{\Theta}_\sat)\,p(\bs{\Theta}_\Phi)
%\end{multline}
\begin{multline}
	p(\{\D^{(i)}\}, \D_\sat \given \{\bs{\Theta}^{(i)}\}, \bs{\Theta}_\sat, \bs{\Theta}_\Phi ) = \\
		p(\D_\sat \given \W_\sat) \, \prod_i \, p(\D^{(i)} \given \W^{(i)}) \,
			p(\bs{\Theta}^{(i)} \given \bs{\Theta}_\sat, \bs{\Theta}_\Phi)
\end{multline}
All parameters are summarized in Table~\ref{tbl:params}.
